Table-Driven Customization of Web Applications

### 2020-05-19 paper.tex
skeleton coming together

Many software applications don't meet the precise needs of their users.
When this happens, the only recourse is to complain to the developers,
or, more likely, to simply give up. Back in 1977, in Personal Dynamic
Media \citep{kay1977}, Alan Kay envisioned personal computing as a
medium that let a user ``mold and channel its power to his own needs,''
but today, software behaves more like concrete than clay.

\begin{itemize}
\tightlist
\item
  many approaches to customizing software. Most are imperative command
  based. Can't get around it.
\item
  there are successful systems for visual SQL queries: Airtable,
  Sieuferd. Even spreadsheets.
\item
  what if you could see the data queries powering a website and directly
  modify them yourself?
\item
  we call this table-driven customization.
\end{itemize}

To make it work with the real world, we develop Wildcard, shim this on
top of existing web apps. Show in Section Examples that it really works!
can do tons of useful stuff with real apps.

Architecture: Describe implementation of Wildcard. We've developed
generic abstractions that make this system even more extensible in the
future. we also propose a future architecture where apps are built for
this; they expose their data queries and their structured data results.

Design Principles: explain ideas behind the system, generalizable to
software customization more generally

Related Work: table editors,

\hypertarget{examples}{%
\section{Examples}\label{examples}}

@@

\hypertarget{system-architecture}{%
\section{System architecture}\label{system-architecture}}

@@

\hypertarget{design-aspects}{%
\section{Design aspects}\label{design-aspects}}

@@

\hypertarget{proximity-to-use}{%
\subsection{Proximity to use}\label{proximity-to-use}}

@@

\hypertarget{related-work}{%
\section{Related Work}\label{related-work}}

@@

  \begin{itemize}
  \tightlist
  \item
    fundamentally rearchitected around table adapters
  \item
    evaluated on many more websites
  \item
    more fully describing how the system works
  \end{itemize}
\item
  browser extensions
\item
  instrumental interaction, Scotty, Webstrates
\item
  customization research tools: Chickenfoot, Coscripter

  \begin{itemize}
  \tightlist
  \item
    Wildcard, like Chickenfoot, wants to hide HTML from users. But we
    show a structured data view, whereas Chickenfoot shows nothing
  \end{itemize}
\item
  desktop customization: Applescript, VBA, COM
\item
  browser dev tools
\item
  database GUIs
\item
  spreadsheets
\end{itemize}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
%% \bibliographystyle{ACM-Reference-Format}
%%\bibliography{references2.bib}

>>>>

There have been many attempts at empowering end users to customize their
software by offering them a simplified programming tool. Some scripting
languages (AppleScript, Chickenfoot) have a friendly syntax that
resembles natural language. Visual programming tools (Mac Automator,
Zapier) eliminate text syntax entirely. Macro recorders (Applescript,
Helena, WebVCR) remove some of the initial programming burden by letting
a user start by concrete demonstrations.

These approaches have many differences, but they all share something in
common: an imperative programming model, with mutable variables,
conditionals and loops. End users ultimately must use these traditional
programming constructs to express their ideas, and the object of
interest is fundamentally a \emph{script}, a sequence of commands.

For decades, we have known about an easier model: \emph{direct
manipulation} \citep{shneiderman1983}, where ``visibility of the object
of interest'' replaces ``complex command language syntax''. Direct
manipulation is widely used in GUIs today, but rarely applied for
software customization. In this work, we ask: what would it look like to
build a software customization interface that deeply relies on direct
manipulation?

We take particular inspiration from spreadsheets and visual database
query interfaces \citep{2020a, bakke2016}. These tools enable users to
construct queries, edit data, and compute derived values through direct
manipulation of the data itself, with intermediate feedback available at
every step. These tools have successfully enabled millions of end users
to view, edit, and compute with data.

We have developed a technique called \emph{table-driven customization}
which applies ideas from visual query interfaces in the context of
software customization. We provide the end user with a table view where
they can directly manipulate the internal state of an application.
Changes in the structured table view result in changes to the original
user interface of the application, enabling the user to customize an
application with immediate feedback.

To demonstrate the viability of this approach, we have developed a
browser extension called Wildcard which implements table-driven
customization in the context of the Web. Wildcard works with existing
web applications by using web scraping techniques to extract relevant
application state into a table and to reify changes back into the
original application UI. In Section~\ref{sec:examples} we present
examples of using Wildcard to implement useful customizations across
many real websites, ranging from sorting lists of data to adding whole
new features to applications.

In Section~\ref{sec:architecture} we explain the architecture of the
Wildcard system, focusing on the key abstraction of the \emph{table
adapter} which manages a single table of data. We explain why some of
these abstractions are general enough to support many customizations
beyond the ones we've built so far in Wildcard---for example, we explore
how applications themselves could be re-architected to expose tabular
data to end users, rather than relying on web scraping techniques.

Table-driven customization embodies several design principles for
building customizable software systems, described in
Section~\ref{sec:design-principles}:

\begin{itemize}
\tightlist
\item
  \emph{Data-oriented programming}: End users should be able to
  customize an application by examining and modifying its data, rather
  than by issuing imperative commands to an API
\item
  \emph{Black-box + semantic}: Typically, ``black-box'' customization
  tools, which don't rely on official extension APIs, resort to offering
  low-level APIs for customization. Instead, we suggest a third-party
  community of semantic wrappers, offering end users a more convenient
  semantic API for customization, even if it wasn't built in by the
  original developer.
\item
  \emph{Gentle slope}: Customization tools should offer a ``gentle
  slope'' from normal use to customization. We explore how our approach
  offers end users opportunities to start customizing software with
  almost no effort, which we hope can make customization a viable part
  of everyday software usage rather than a rare event.
\end{itemize}

Table-driven customization relates to existing work in many areas. Our
goals overlap with many software customization tools, in the browser and
on the desktop. Our methods overlap with direct manipulation interfaces
for working with data, including visual database query systems and
spreadsheets. We explore these connections and more in
Section~\ref{sec:related-work}.

\hypertarget{sec:examples}{%
\section{Examples}\label{sec:examples}}

@@

\hypertarget{sec:architecture}{%
\section{System architecture}\label{sec:architecture}}

@@

\hypertarget{sec:design-principles}{%
\section{Design principles}\label{sec:design-principles}}

@@

\hypertarget{gentle-slope}{%
\subsection{Gentle slope}\label{gentle-slope}}

@@

\hypertarget{sec:related-work}{%
\section{Related Work}\label{sec:related-work}}

@@

  \begin{itemize}
  \tightlist
  \item
    fundamentally rearchitected around table adapters
  \item
    evaluated on many more websites
  \item
    more fully describing how the system works
  \end{itemize}
\item
  database GUIs:

  \begin{itemize}
  \tightlist
  \item
    Sieuferd \citep{bakke2016}
  \item
    Airtable \citep{2020a}
  \item
    other similar tools?
  \item
    Liu \& Jagadish: rules for a spreadsheet algebra for database
    queries \citep{liu2009}
  \item
    mention DIRECT MANIPULATION
  \end{itemize}
\item
  browser extensions
\item
  instrumental interaction, Scotty \citep{eagan2011}, Webstrates
\item
  customization research tools: Chickenfoot, Coscripter

  \begin{itemize}
  \tightlist
  \item
    Wildcard, like Chickenfoot, wants to hide HTML from users. But we
    show a structured data view, whereas Chickenfoot shows nothing
  \end{itemize}
\item
  visual customization tools:

  \begin{itemize}
  \tightlist
  \item
    Mac Automator
  \item
    Zapier
  \item
    IFTTT
  \item
    These are weakly direct manipulation. But usually the thing you
    directly manipulate in these UIs is the \emph{script}: not the
    actual \emph{data} you want to operate on. VERY DIFFERENT.
  \end{itemize}
\item
  desktop customization: Applescript, VBA, COM
\item
  browser dev tools
\item
  database GUIs
\item
  spreadsheets
\item
  ScrAPIr
\end{itemize}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{wildcard-onward-bibtex.bib}
